---
title: "p8105_hw3_fl2714"
author: "Fangchi"
date: "2024-10-10"
output: github_document
---

```{r}

library(tidyverse)
library(readxl)
library(janitor)
library(lubridate)
library(knitr)
library(dplyr)

```

## Problem 2 - NHANES Accelerometer Data

```{r}

demographics_df <- 
  read_csv("data/nhanes_covar.csv",skip = 4) |> 
  janitor::clean_names() |>
  filter(age >= 21, !is.na(education), !is.na(sex)) 

accelerometer_df <- 
  read_csv("data/nhanes_accel.csv") |>
  janitor::clean_names()

```

```{r}

# Join the demographic and accelerometer datasets by participant ID
final_data <- 
  left_join(accelerometer_df, demographics_df, by = "seqn") |>
  drop_na()

```

```{r}

# Encode variables with appropriate data types
final_data <- final_data %>%
  mutate(
    education = factor(education, levels = 1:3, labels = c("Less than High School", "High School", "College")),
    sex = factor(sex, levels = c(1, 2), labels = c("Male", "Female"))
  )

```

```{r}
# Table of men and women in each education category
education_sex_table <- final_data %>%
  group_by(education, sex) %>%
  summarise(count = n(), .groups = "drop") %>%
  pivot_wider(names_from = sex, values_from = count) %>%
  mutate(Total = Male + Female)

# Display the table using knitr
knitr::kable(education_sex_table, caption = "Number of Men and Women in Each Education Category")

```



```{r}

# Plot age distribution by sex and education level
ggplot(final_data, aes(x = age, fill = sex)) +
  geom_density(alpha = 0.5) +
  facet_wrap(~ education) +
  labs(title = "Age Distribution by Sex and Education Level",
       x = "Age", y = "Density",
       fill = "Sex") +
  theme_minimal()

```
### Observations

- **Age Distribution**: Age distributions vary across education levels, with a higher concentration of younger participants in the **College** group and older participants in the **Less than High School** and **High School** groups.
- **Sex Differences**: For **College**-educated individuals, there is a slight female predominance across most age ranges, while the **Less than High School** and **High School** groups show a more balanced or slightly male-dominant distribution.
- **Peak Ages**: The **College** group has a noticeable peak in the 20–30 age range, reflecting a younger demographic, whereas **Less than High School** and **High School** groups have peaks around 60 years, indicating an older demographic for those with lower education levels.



```{r}

# Calculate total daily activity for each participant
final_data <- final_data %>%
  mutate(total_activity = rowSums(select(final_data, starts_with("min"))))

# Plot total activity vs. age by sex and education level
ggplot(final_data, aes(x = age, y = total_activity, color = sex)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "loess", se = FALSE) +
  facet_wrap(~ education) +
  labs(title = "Total Daily Activity vs Age by Sex and Education Level",
       x = "Age", y = "Total Daily Activity",
       color = "Sex") +
  theme_minimal()

```

### Observations

- **Age and Total Activity**: Total activity levels vary by age, with a tendency to decrease among older participants.
- **Sex Differences**: Men generally show slightly higher activity levels than women across most age groups.
- **Education Levels**: The patterns of activity by age are similar across education levels, though college-educated participants show slightly higher activity overall, possibly reflecting lifestyle differences.
- **Trend**: The smooth trend line illustrates a gradual decline in activity with age for both men and women.

```{r}

# Reshape to long format for minute-by-minute activity
activity_long <- final_data %>%
  pivot_longer(cols = starts_with("min"), 
               names_to = "minute", 
               values_to = "activity") %>%
  mutate(minute = as.numeric(str_remove(minute, "min")))

# Plot 24-hour activity patterns
ggplot(activity_long, aes(x = minute, y = activity, color = sex)) +
  geom_smooth(se = FALSE) +
  facet_wrap(~ education) +
  scale_x_continuous(breaks = seq(0, 1440, 240),
                     labels = c("12AM", "4AM", "8AM", "12PM", "4PM", "8PM", "12AM")) +
  labs(title = "24-Hour Activity Time Courses by Education Level and Sex",
       x = "Time of Day", y = "Activity (MIMS)",
       color = "Sex") +
  theme_minimal()

```

### Observations

- **Daily Activity Peaks**: All show two main peaks in activity — one in the morning around 8 AM and another in the evening around 6 PM.

- **Differences by Education Level**:  
  - **Less than High School** and **College** groups have a more pronounced morning peak than the **High School** group. The **College** group maintains slightly higher activity levels throughout the day, while the **Less than High School** and **High School** groups have lower midday activity.

- **Sex Differences**:  
  - Across all education levels, women show slightly higher morning activity peaks than men.

## Problem 3 - Citi Bike

```{r}

Jan_2020_df <- read_csv("data/citibike/Jan 2020 Citi.csv", show_col_types = FALSE) %>%
  janitor::clean_names() %>%
  mutate(year = 2020, month = "January")  

Jul_2020_df <- read_csv("data/citibike/July 2020 Citi.csv", show_col_types = FALSE) %>%
  janitor::clean_names() %>%
  mutate(year = 2020, month = "July")

Jan_2024_df <- read_csv("data/citibike/Jan 2024 Citi.csv", show_col_types = FALSE) %>%
  janitor::clean_names() %>%
  mutate(year = 2024, month = "January")

Jul_2024_df <- read_csv("data/citibike/July 2024 Citi.csv", show_col_types = FALSE) %>%
  janitor::clean_names() %>%
  mutate(year = 2024, month = "July")

# merge all data
all_months_df <- bind_rows(Jan_2020_df, Jul_2020_df, Jan_2024_df, Jul_2024_df)

# summarize 
ride_summary <- all_months_df %>%
  group_by(year, month, member_casual) %>%
  summarise(total_rides = n(), .groups = 'drop') %>%
  pivot_wider(names_from = member_casual, values_from = total_rides, values_fill = 0)

# view results
kable(ride_summary, caption = "Total Number of Rides by Year, Month, and Rider Type (Casual and Member)")

```

### Observations

- **Increase in Total Rides**:  
  Both **casual** and **member** rides increased significantly from 2020 to 2024.

- **Seasonal Variation**:  
  For both years, **July** has a much higher ride count than **January**. 

- **Casual vs. Member Rides**:  
  - **Members** consistently have higher ride counts than casual users in each month and year. The number of rides from memeber continuously increased.
  - The increase in casual rides from **January** to **July** is more pronounced, which suggests that casual users may be more motivated by leisure or seasonal factors, while members have a relatively stable and predictable usage pattern.

```{r}
top_stations_jul2024 <- Jul_2024_df %>%
  group_by(start_station_name) %>%
  summarise(total_rides = n(), .groups = 'drop') %>%
  arrange(desc(total_rides)) %>%
  slice_head(n = 5)  

kable(top_stations_jul2024, 
      col.names = c("Starting Station", "Number of Rides"), 
      caption = "Top 5 Most Popular Starting Stations in July 2024")
```




```{r}

median_duration_df <- all_months_df %>%
  group_by(year, month, weekdays) %>%
  summarise(median_duration = median(duration, na.rm = TRUE), .groups = 'drop')

ggplot(median_duration_df, aes(x = weekdays, y = median_duration, color = factor(year))) +
  geom_line(aes(group = year), size = 1) +
  facet_wrap(~ month, ncol = 3) +
  labs(title = "Median Ride Duration by Day of Week, Month, and Year",
       x = "Day of the Week",
       y = "Median Ride Duration (Minutes)",
       color = "Year") +
  theme_minimal() +
  theme(
    legend.position = "bottom",
    axis.text.x = element_text(angle = 45, hjust = 1)  
  )

```

### Observations

- **Seasonal Variation**:  
  Ride durations are generally higher in **July** than in **January** for both 2020 and 2024, likely due to increased recreational use during summer.

- **Yearly Differences**:  
  In 2020, **July** shows a sharp peak in ride duration on weekends, exceeding 15 minutes. In 2024, weekend durations are still higher but less pronounced, suggesting a shift towards shorter trips.

- **Weekday Patterns**:  
  Weekdays consistently show shorter ride durations, likely reflecting commuter usage, while weekends have longer durations, indicating more recreational rides.

```{r}

data_2024 <- all_months_df %>%
  filter(year == 2024)

ggplot(data_2024, aes(x = member_casual, y = duration, fill = rideable_type)) +
  geom_violin(alpha = 0.7) +
  facet_wrap(~ month, ncol = 2) +
  scale_y_continuous(limits = c(0, quantile(data_2024$duration, 0.95)),  
                     name = "Ride Duration (Minutes)") +
  labs(
    title = "Distribution of Ride Duration by Membership, Bike Type, and Month (2024)",
    x = "Membership Status",
    fill = "Bike Type"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "bottom"
  )

```
### Observations

- **Electric vs. Classic Bikes**:  
  Electric bikes tend to have shorter ride durations than classic bikes, likely due to faster speeds allowing quicker trips.

- **Membership Differences**:  
  Casual riders show longer ride durations, especially on classic bikes, indicating more recreational use. Members generally have shorter, more consistent durations, suggesting commuting use.

- **Seasonal Variation**:  
  In **July**, ride durations are longer for both casual and member users, reflecting increased recreational use in warmer months.

